---
title: "Accessing Bottom Trawl Survey Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accessing Bottom Trawl Survey Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  eval = FALSE
)
```

## NEFSC Bottom Trawl Survey Data in {survdat}

The R package {survdat} allows you to pull and process NEFSC bottom trawl survey data. More information on the {survdat} package can be found here: https://noaa-edab.github.io/survdat/index.html

Before using this package, you will need to submit a number of ITD tickets:
- Install of Oracle Instant Client 
- Creation of Oracle username and password. Also request access to NEFSCDB1 and NEFSC_USERS databases in this ticket. 
- Install of {ROracle} package

You must also be behind the NEFSC firewall (remote using VPN) for R to connect to Oracle databases. 

## Installation

To install {survdat} directly from Github:
```{r}
remotes::install_github("NOAA-EDAB/survdat",build_vignettes = TRUE)
```

The package {dbutils} must also be installed for R to connect to Oracle databases. More info on {dbutils} here: https://andybeet.github.io/dbutils/
```{r}
remotes::install_github("andybeet/dbutils")
```

To connect to Oracle databases and pull survdat data
```{r}
channel <- dbutils::connect_to_database("server name","your username")
survdat::get_survdat_data(channel)
```

The function get_survdat_data() has parameters to include conversion factors (conversion.factor), species biological data (getBio), and length data (getLengths) which can be set = TRUE or FALSE.

## POC for Troubleshooting

Contact Andy Beet (andrew.beet@noaa.gov) for issues with {dbutils} and {survdat} package help and data pulls.

If you do not have access to all the required tables in SVDBS to run {survdat} functions, the function 'get_survdat_data' will result in the following error:

"Error in .oci.GetQuery(conn, statment, data = data, prefetch = prefetch,...) ORA-00942: table or view does not exist"

If this occurs, contact Jakub Kircun with PEMAD (jakub.cirkun@noaa.gov) for SVDBS table access.
