---
title: "Using MRIP Data for Recreational Indicators"
author: "Stephanie Owen"
vignette: >
  %\VignetteIndexEntry{Using MRIP Data for Recreational Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

## Obtaining Raw MRIP Data

Data pulls from NOAA Fisheries' Marine Recreational Information Program (MRIP) can be done through the MRIP Query Tool (https://www.fisheries.noaa.gov/data-tools/recreational-fisheries-statistics-queries). 
 
Queries for recreational trips, catch, and landings can be performed using the functions 'get_mrip_catch' and 'get_mrip_trips' located in 'R/get_mrip_data.R' in this package.

This vignette shows example data pulls and plots used for Black Sea Bass:
 - Total Recreational Catch
 - Total Recreational Trips
 - Recreational Effort (proportion of trips targeting black sea bass)
 - Total Recreational Landings
 
## Run MRIP functions and plot time series

### Black Sea Bass Recreational Catch

```{r}
bsb_catch <- get_mrip_catch(species = "BLACK SEA BASS", type = "all")
catch <- bsb_catch$data

head(catch)

plt_indicator(catch)
```

### Black Sea Bass Recreational Landings

```{r}
bsb_landings <- get_mrip_catch(species = "BLACK SEA BASS", type = "landings")
landings <- bsb_landings$data

head(landings)

plt_indicator(landings)
```

### Black Sea Bass Total Recreational Trips

Recreational trip queries must be done by year (single query for each year). The earliest possible year is 1981. 
Additionally, each region must be queried separately (i.e. Mid-Atlantic and North Atlantic, or by state).

The following queries Black Sea Bass trips for the North Atlantic and Mid-Atlantic from 2020-2024.

```{r}
rec_trips <- create_rec_trips(files = files)

plt_indicator(rec_trips)
```

