---
title: "Using EDAB Utilities Functions for NE ESPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using EDAB Utilities Functions for NE ESPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  eval = FALSE
)
```


## EDAB Utilities 

This vignette demonstrates use of functions within the EDABUtilities package (https://github.com/NEFSC/READ_EDAB_Utilities/tree/main) in examining ecosystem data for ESPs. The example functions here apply input data (as .nc files) and shape files utilized in the golden tilefish ESP. 

### Crop netCDFs of bottom temperature and bottom salinity to shape file of BTS strata

Crops a 2D netCDF based on the extent of a shape file.

```{r}
crop_bottomT <- EDABUtilities::crop_nc_2d(data.in = c(here::here('data-raw','glorys_bottomT.nc')),
                            output.files = c(here::here('data-raw','crop_glorys_bottomT.nc')),
                            shp.file = here::here('data-raw/shapefiles','BTS_STRATA.shp'),
                            #area.names = c('01030'),
                            var.name = 'bottomT',
                            write.out = F)

terra::plot(crop_bottomT[[1]])
```

```{r}
crop_bottomS <- EDABUtilities::crop_nc_2d(data.in = c(here::here('data-raw','glorys_bottomS.nc')),
                            output.files = c(here::here('data-raw','crop_glorys_bottomS.nc')),
                            shp.file = here::here('data-raw/shapefiles','BTS_STRATA.shp'),
                            var.name = 'bottomS',
                            write.out = F)

terra::plot(crop_bottomS[[1]])

```


### Make 2d gridded time series: calculate number of days (nd) above a bottom temperature threshold (14C) in the golden tilefish BTS strata

Provides a data frame of degree-day family statistics

```{r}
dd <- EDABUtilities::make_2d_deg_day_ts(data.in = c(here::here('data-raw','glorys_bottomT.nc')),
                              output.files = c(here::here('data-raw','GLORYS_monthly_bottomT.nc')),
                              shp.file = here::here('data-raw/shapefiles','BTS_STRATA.shp'),
                              var.name = 'bottomT',
                              area.names = c('01030', '01040', '01070', '01080', '01110', '01120', 
                                            '01140', '01150', '01670', '01680', '01710', '01720', 
                                            '01750', '01760'),
                   ref.value = 14,
                   type = 'above',
                   statistic = 'nd',
                   write.out = F)

dd_df = dplyr::bind_rows(dd)

ggplot2::ggplot(dd_df,ggplot2::aes(x = ls.id, y = value, fill = area))+
  ggplot2::geom_bar(stat = 'identity', position = ggplot2::position_dodge(width = 1.0))+
  ggplot2::labs(title = 'Number of days above 14C', fill = "Strata Number") +
  ggplot2::xlab('GTF Strata') +
  ggplot2::ylab('Days') +
  ggplot2::theme_bw()
```

