---
title: "Using {survdat} for Bottom Trawl Indicators"
author: "Stephanie Owen"
vignette: >
  %\VignetteIndexEntry{Using survdat for Bottom Trawl Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

## Bottom Trawl Indicators in NEESP2

This package contains functions for producing bottom trawl indicators from {survdat} data pulls: swept area biomass, stratified mean biomass, species condition, diet, and range. This vignette provides example output data tables and plots using these functions. 

For more information on accessing {survdat} data, see vignette "Accessing Bottom Trawl Survey Data".

### Species Condition - Black sea bass

Uses 'species_condition' and 'plot_condition' functions in R/create_condition_indicator.R and R/plot_condition.R

```{r}
condition <- species_condition(data = NEesp2::survdat_subset, 
                               LWparams = LWparams, 
                               species.codes = species.codes, 
                               output = "esp") 
bsb_condition <- condition %>%
  subset(Species == 'Black sea bass')
head(bsb_condition)
```

```{r}
plot_condition(data = bsb_condition, var = 'Black sea bass') 
```

### Species Range - Black sea bass in fall

Uses 'species_range' function in R/create_bottomtrawl_indicators.R

```{r}
range <- species_range(data = NEesp2::survdat_subset, species = NEesp2::species)
bsb_range <- range %>%
  subset(SPECIES == 'BLACK SEA BASS') %>%
  subset(SEASON == 'FALL') %>%
  subset(INDICATOR_NAME == c('range_lat','range_lon'))
head(bsb_range)
```

```{r}
bsb_range %>%
  ggplot2::ggplot(ggplot2::aes(x = YEAR,
                               y = DATA_VALUE,
                               color = INDICATOR_NAME,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() + 
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Black Sea Bass Range in Fall (Lat + Lon)", 
                x = "Year",
                y = "Range (degrees)") +
  ggplot2::facet_wrap('INDICATOR_NAME')
```

### Swept area biomass - Black sea bass in fall

Uses 'create_swept_area' indicator in R/create_bottomtrawl_indicators.R 

Modified from 'calc_swept_area' in {survdat} package.

```{r}
bsb_swept_area <- create_swept_area(
  surveyData = NEesp2::survdat_subset,
  areaPolygon = "NEFSC strata",
  areaDescription = "STRATA",
  filterByArea = "all",
  filterBySeason = "FALL",
  groupDescription = "SVSPP",
  filterByGroup = "all",
  mergesexFlag = T,
  tidy = F,
  q = NULL,
  a = 0.0384
) %>%
  dplyr::filter(SVSPP == "141")

head(bsb_swept_area)
```

```{r}
bsb_swept_area %>%
  ggplot2::ggplot(ggplot2::aes(x = YEAR,
                               y = DATA_VALUE,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() + 
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Black Sea Bass Swept Area Biomass (Fall)", 
                x = "Year",
                y = "Swept Area Biomass") 
```


### Stratified mean biomass - Black sea bass in fall

Uses 'create_stratified_mean' indicator in R/create_bottomtrawl_indicators.R 

Modified from 'calc_stratified_mean' in {survdat} package.

```{r}
bsb_strat_mean <- create_stratified_mean(
  surveyData = survdat_subset,
  areaPolygon = "NEFSC strata",
  areaDescription = "STRATA",
  filterByArea = "all",
  filterBySeason = "FALL",
  groupDescription = "SVSPP",
  filterByGroup = "all",
  mergesexFlag = T,
  tidy = F,
  returnPrepData = F
) %>%
  dplyr::filter(SVSPP == "141")

head(bsb_strat_mean)
```

```{r}
bsb_strat_mean %>%
  ggplot2::ggplot(ggplot2::aes(x = YEAR,
                               y = DATA_VALUE,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() + 
  ggplot2::theme_bw() +
  ggplot2::labs(title = "Black Sea Bass Stratified Mean Biomass (Fall)", 
                x = "Year",
                y = "Stratified Mean Biomass") 
```

