---
title: "Using {survdat} for Bottom Trawl Indicators"
author: "Stephanie Owen and Abigail Tyrell"
vignette: >
  %\VignetteIndexEntry{Using survdat for Bottom Trawl Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

## Bottom Trawl Indicators in NEESP2

This package contains functions for producing bottom trawl indicators from {survdat} data pulls: swept area biomass, stratified mean biomass, species condition, diet, and range. This vignette provides example output data tables and plots using these functions. 

For more information on accessing {survdat} data, see vignette "Accessing Bottom Trawl Survey Data".

### Species Condition - Black sea bass

#### Create and plot species condition

Uses 'species_condition' and 'plot_condition' functions in R/create_condition_indicator.R and R/plot_condition.R

```{r}
condition <- species_condition(
  data = NEesp2::survdat_subset,
  LWparams = LWparams,
  species.codes = species.codes,
  output = "esp"
)
bsb_condition <- condition %>%
  subset(Species == "Black sea bass")
head(bsb_condition)
```

```{r}
plot_condition(data = bsb_condition, var = "Black sea bass")
```

Data are aggregated at the EPU level by default, but the data can also be assessed as a unit.

```{r}
condition <- species_condition(
  data = NEesp2::survdat_subset,
  LWparams = LWparams,
  species.codes = species.codes,
  output = "esp",
  by_EPU = FALSE
)
bsb_condition <- condition %>%
  subset(Species == "Black sea bass")
head(bsb_condition)
```

```{r}
plot_condition(data = bsb_condition, var = "Black sea bass")
```

#### Create species condition for the State of the Ecosystem report

By setting `output = "soe"`, the `species_condition` function will format the output in the same style as `ecodata::condition`. Note that black sea bass is the only species included in the sample dataset shown below, but if additional species were present in the data, the output would be aggregated at the species level.

```{r}
condition <- species_condition(
  data = NEesp2::survdat_subset,
  LWparams = LWparams,
  species.codes = species.codes,
  output = "soe"
)
head(condition)
```

#### Additional condition queries

`species_condition` can also assess the data by sex and by length bin. If these options are selected, the output will be a "full" dataset that contains additional data columns identifying the grouping parameters and data variance. For example:

```{r}
condition <- species_condition(
  data = NEesp2::survdat_subset,
  LWparams = LWparams,
  species.codes = species.codes,
  by_sex = TRUE,
  length_break = c(0, 20, 65)
)
head(condition)
```


### Species Range - Black sea bass in fall

Uses 'species_range' function in R/create_bottomtrawl_indicators.R

```{r}
range <- species_range(data = NEesp2::survdat_subset, species = NEesp2::species)
bsb_range <- range %>%
  subset(SPECIES == "BLACK SEA BASS") %>%
  subset(SEASON == "FALL") %>%
  subset(INDICATOR_NAME == c("range_lat", "range_lon"))
head(bsb_range)
```

```{r}
bsb_range %>%
  ggplot2::ggplot(ggplot2::aes(
    x = YEAR,
    y = DATA_VALUE,
    color = INDICATOR_NAME,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    title = "Black Sea Bass Range in Fall (Lat + Lon)",
    x = "Year",
    y = "Range (degrees)"
  ) +
  ggplot2::facet_wrap("INDICATOR_NAME")
```

### Swept area biomass - Black sea bass in fall

Uses 'create_swept_area' indicator in R/create_bottomtrawl_indicators.R 

Modified from 'calc_swept_area' in {survdat} package.

```{r}
bsb_swept_area <- create_swept_area(
  surveyData = NEesp2::survdat_subset,
  areaPolygon = "NEFSC strata",
  areaDescription = "STRATA",
  filterByArea = "all",
  filterBySeason = "FALL",
  groupDescription = "SVSPP",
  filterByGroup = "all",
  mergesexFlag = T,
  tidy = F,
  q = NULL,
  a = 0.0384
) %>%
  dplyr::filter(SVSPP == "141")

head(bsb_swept_area)
```

```{r}
bsb_swept_area %>%
  ggplot2::ggplot(ggplot2::aes(
    x = YEAR,
    y = DATA_VALUE,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    title = "Black Sea Bass Swept Area Biomass (Fall)",
    x = "Year",
    y = "Swept Area Biomass"
  )
```


### Stratified mean biomass - Black sea bass in fall

Uses 'create_stratified_mean' indicator in R/create_bottomtrawl_indicators.R 

Modified from 'calc_stratified_mean' in {survdat} package.

```{r}
bsb_strat_mean <- create_stratified_mean(
  surveyData = survdat_subset,
  areaPolygon = "NEFSC strata",
  areaDescription = "STRATA",
  filterByArea = "all",
  filterBySeason = "FALL",
  groupDescription = "SVSPP",
  filterByGroup = "all",
  mergesexFlag = T,
  tidy = F,
  returnPrepData = F
) %>%
  dplyr::filter(SVSPP == "141")

head(bsb_strat_mean)
```

```{r}
bsb_strat_mean %>%
  ggplot2::ggplot(ggplot2::aes(
    x = YEAR,
    y = DATA_VALUE,
  )) +
  ggplot2::geom_point() +
  ggplot2::geom_path() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    title = "Black Sea Bass Stratified Mean Biomass (Fall)",
    x = "Year",
    y = "Stratified Mean Biomass"
  )
```

