---
title: "Using MRIP Data for Recreational Indicators"
author: "Stephanie Owen"
vignette: >
  %\VignetteIndexEntry{Using MRIP Data for Recreational Indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all()
```

## Obtaining Raw MRIP Data

Data pulls from NOAA Fisheries' Marine Recreational Information Program (MRIP) can be done through the MRIP Query Tool (https://www.fisheries.noaa.gov/data-tools/recreational-fisheries-statistics-queries). 
 
At this time, queries must be saved as a csv and loaded into R. The raw data includes lines of metadata and description and does not format nicely. The following lines of code remove the extra information and modifies the remaining data into columns while reading in each csv file. The files used in this vignette are found in the 'data-raw' folder in this package. 

Catch, landings, and effort data can be read in as one file with multiple years. Directed trips must be downloaded for each year of interest. 

```{r}
# catch data
mrip_catch <- read.csv(here::here('data-raw/mrip_BLACK_SEA_BASS_catch_series.csv'),
                                  skip = 46, # of rows you want to ignore
                                  na.strings = ".") |>
  janitor::clean_names(case = "all_caps")

# landings data
mrip_landing <- read.csv(here::here('data-raw/mrip_BLACK_SEA_BASS_harvest.csv'),
                         skip = 46, # of rows you want to ignore
                         na.strings = ".") |>
  janitor::clean_names(case = "all_caps")

# effort data 
mrip_effort <- read.csv(here::here('data-raw/mrip_effort_series.csv'),
                        skip = 44, # of rows you want to ignore
                        na.strings = ".") |>
  janitor::clean_names(case = "all_caps")

# trips files
path = (here::here('data-raw/mrip_directed_trips'))

files <- list.files(path = path, 
                    pattern = utils::glob2rx('mrip*.csv'),
                    full.names = TRUE)
```

This vignette shows example data pulls and plots used for the 2025 black sea bass ESP report card:
 - Total Recreational Catch
 - Total Recreational Trips
 - Recreational Effort (proportion of trips targeting black sea bass)
 - Total Recreational Landings
 
## Run MRIP functions and plot time series

The code for functions referenced here can be found in 'create_recreational_indicators.R' and 'plot_indicator_ts.R'.

### Black Sea Bass Recreational Catch

Query 'Time Series' under 'Catch Data'.

Choose years of interest, summarize by Annual, Calendar Year, Atlantic coast by state, Black Sea Bass, all modes and areas, Total Catch

Download csv as output

```{r}
total_rec_catch <- create_total_rec_catch(mrip_catch)

plt_indicator(total_rec_catch)
```

### Black Sea Bass Recreational Landings

Query 'Time Series' under 'Catch Data'.

Choose years of interest, summarize by Annual, Calendar Year, Atlantic Coast by State, Black Sea Bass, all modes and areas, Harvest (A + B1), choose # of fish/weight (lbs), mean length and weight

Download csv as output
```{r}
total_rec_landings <- create_total_rec_landings(mrip_landing)

plt_indicator(total_rec_landings)
```

### Black Sea Bass Total Recreational Trips

Query 'Directed Trip' under 'Effort Data'.

Choose year of interest, summarize by Annual, Calendar Year, Atlantic coast by state, Black Sea Bass, all modes and areas, Primary Target
 
Download csv as output

```{r}
rec_trips <- create_rec_trips(files = files)

plt_indicator(rec_trips)
```

### Black Sea Bass Recreational Effort

Note: must run 'create_rec_trips' prior to obtain 'rec_trips' dataset.

Query 'Time Series' under 'Effort Data'.

Choose years of interest, summarize by Annual, Calendar Year, Atlantic Coast by State, all modes and areas

Download csv as output
```{r}
prop_sp_trips <- create_prop_sp_trips(total_trips = mrip_effort, 
                                      species_trips = rec_trips)

plt_indicator(prop_sp_trips)
```

